Automatic word stress annotation of Russian running text (max 8 pages)

### 2015-01-16 rusaccent-nodalida-2015/rusaccent.tex
More conclusions and some more bibtex.
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58375 72bbbca6-d526-0410-a7d9-f06f51895060

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 
1998), but very few studies have targeted stress-marking itself, and they tend to
focus on placing stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{compute the difference in stress
ambiguity between wordform lists and running text, i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

@@

State-of-the-art morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} To our knowledge,
no open-source resources are available for analyzing and generating stressed wordforms.
Therefore, we developed free and open-source finite-state tools capable of analyzing and
generating stressed wordforms, based on the well-known 
\emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}. A finite-state 
transducer\footnote{Using two-level 
morphology, xfst/hfst, blah, blah, blah} (FST) generates all possible morphosyntactic readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

@@

One of the main points of this paper is to highlight the importance of 
syntactic context in the Russian wordstress placement task. \cite{hall_sproat_russian_2013} 
achieve 98.7\% accuracy on stress placement of individual wordforms. To 
understand this result, it should be remembered that Russian has a considerable
number of words that are ambiguous with regard to stress. Based on the surface 
forms in our FST (which is based on the same dictionary used for 
\cite{hall_sproat_russian_2013}), we calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions.} 
Compare this with the finding of our experiment, that in running 
text, at least 7.5\% of the tokens are stress-ambiguous. This is because many of 
the lexemes with shifting stress patterns are high-frequency. Any system that
cannot leverage syntactic context to disambiguate these wordforms cannot 
achieve accuracy above 92.5\% without making an higher number of errors.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.

We have shown that our Russian Constraint Grammar is able to

Limitations

Another relevant issue is whether the ambiguity can be resolved at all. For example, when is stress marked in 
authentic by-native-for-native texts? A computer should not be expected to predict the stress on 
such words (although a computer could possibly identify words that should be marked for 
natives.)

Future work:
Two sections: known words and unknown words
\begin{itemize}
\item New domains and reading levels
\item As with any lexicon- and rule-based system, those components can be expanded and improved, especially considering the short span over which they were developed.
\item Integrate a more principled unknown word-guessing method (homicevic)
\item Compare and/or integrate with stochastic methods. So far stochastic
methods have only been used for guessing stress on unknown words.
\item

>>>>

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. \cite{krivnova_1998}), but very few studies have targeted stress-marking itself, and they tend to
focus on placing stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, \cite{hall_sproat_russian_2013} trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{compute the difference in stress
ambiguity between wordform lists and running text, i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

@@

State-of-the-art morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} To our knowledge,
no open-source resources are available for analyzing and generating stressed wordforms.
Therefore, we developed free and open-source finite-state tools capable of analyzing and
generating stressed wordforms, based on the well-known 
\emph{Grammatical Dictionary of Russian} \cite{Zaliznjak-77}. A Finite-State 
Transducer\footnote{Using two-level 
morphology, xfst/hfst, blah, blah, blah} (FST) generates all possible morphosyntactic readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

@@

One of the main points of this paper is to highlight the importance of 
syntactic context in the Russian wordstress placement task. \cite{hall_sproat_russian_2013} 
achieve 98.7\% accuracy on stress placement of individual wordforms. To 
understand this result, it should be remembered that Russian has a considerable
number of words that are ambiguous with regard to stress. Based on the surface 
forms in our FST (which is based on the same dictionary used for 
\cite{hall_sproat_russian_2013}), we calculate that only 29~518 (1.05\%) 
of the 2~804~492 wordforms contained in our FST are stress-ambiguous. \rr{These 
stress-ambiguous wordforms have an average of 2.01 possible stress positions.} 
Compare this with the finding of our experiment, that in running 
text, at least 7.5\% of the tokens are stress-ambiguous. This is at least partly 
because many of 
the lexemes with shifting stress patterns are high-frequency. Any system that
cannot leverage syntactic context to disambiguate these wordforms cannot 
achieve accuracy above 92.5\% without making an higher number of errors.
Since the task of wordstress placement is virtually always performed on
running text, it seems prudent to make use of surrounding information.
The experiment described in this paper demonstrates that a Constraint Grammar
can effectively improve the accuracy of a stress placement system without 
significantly raising the error rate. Our Russian Constraint Grammar is
under continual development, so we expect higher accuracy in the future.

The experiment was limited by properties of the gold standard corpus, including
its size, genre distribution, and quality. Ideally, a gold standard corpus for Russian wordstress should mark 
whether monosyllabic words are stressed, and of course its annotations should
be accurate. Our gold corpus represents a broad variety of text genres, which 
makes our results more generalizable, but a larger corpus would allow for
evaluating each genre individually. For example, the vast majority of words with
shifting stress are of Slavic origins, so we expect a genre such as technical
writing to have a lower proportion of words with stress ambiguity, since it
uses a more modern lexicon that has not inherited the idiosyncrasies that lead
to wordstress ambiguity.

In addition to genre, it seems likely that text
complexity affects the difficulty of the stress placement task, so subcorpora at
various reading levels would shed light on those interactions as well.

The syllable-guessing algorithm used in this experiment was overly simplistic,
and it was not surprising that it was

Future work/potential applications:
Two sections: known words and unknown words
\begin{itemize}
\item As with any lexicon- and rule-based system, those components can be expanded and improved, especially considering the short span over which they were developed.
\item Integrate a more principled unknown word-guessing method (homicevic)
\item Compare and/or integrate with stochastic methods. So far stochastic
methods have only been used for guessing stress on unknown words.
\item Another relevant issue is whether the ambiguity can be resolved at all. For example, when is stress marked in authentic by-native-for-native texts? A computer should not be expected to predict the stress on such words (although a computer could possibly identify words that should be marked for 
natives.)
